---
title: "Alcohol consumption"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Include the main file in order to do preprocessing.
source("./main.R")
```

### Abstract 

We found out that there is a weak correlation between the poverty index and alcohol consumption. Additionally, alcohol consumption depends on the gender and marital status. Males tend to consume more alcohol than females and those who never married tend to drink more than those separated.

ADD SUMMARY OF PREDICTION

### Introduction

Is alcohol consumption related to factors such as mental health, marital status, income, gender and age?

We did not have a very specific question in mind as we were more curious to analyze the factors related to alcohol consumption in general and find significant or unexpected insights. Some of our initial questions lead us to dead-ends after analyzing the data more in depth, this is why we decided to focus on a broader research, centered around alcohol consumption and with a focus on mental health and personal situations.

One of the initial purpose of developing this project was to find out how mental health relates to alcohol consumption and this has been quite challenging. Establishing this correlation could potentially raise awareness on how alcohol consumption might affect the mental health or that certain untreated mental issues could lead to higher amount consumptions. We were also interested in knowing how alcohol consumption relates to income. For instance, if people of lower income or higher income tend to consume more alcohol. This could potentially raise awareness of vicious cycles in our society. This study could also help verify or deny common beliefs or expectations. For instance, do males consume more alcohol than females? Do married people consume less alcohol than the single and separated/widowed ones? All these could show results that would futher lead researches to continue with more in depth analysis such as why certain relations exist.

### Related Work

##### **Project 1**: [How common is alcohol and drug dependency across the world?](https://ourworldindata.org/alcohol-and-drug-dependency)

This project from the website _Our World in Data_ presents the trends in alcohol consumption and substance use in countries all over the world, with data from 1990 to 2016. The aim is to give a better understanding of the global prevalence of substance use disorders.

Some of the conclusions in the research are:

- Across most countries, the number of alcohol use disorders is higher than other drug use disorders.
- Substance use disorders are more common among men than women.

##### **Project 2**: [SDR, selected alcohol-related causes, per 100

000](https://gateway.euro.who.int/en/indicators/hfa_293-1970-sdr-selected-alcohol-related-causes-per-100-000/)

This project carried out by the World Health Organization provide an interactive graph where we can see the age-standardized death rate per 100000 related to alcohol. The study includes causes such as cancer of esophagus and larynx, alcohol dependence syndrome, chronic liver disease, cirrhosis etc.

##### **Project 3:** [Alcohol as a risk factor for type 2 diabetes](https://care.diabetesjournals.org/content/32/11/2123.full-text.pdf)

This study by the University of Toronto tries to measure the relation between alcohol consumption and the presence of type 2 diabetes. For obtaining the data, the researchers have found several data sources both from public sources or from previous articles.

The main conclusion of this study is that a relationship exists between alcohol consumption and the presence or diabetes in an _u-shaped_ way, both for men and women.

##### **Project 4:** [Risk thresholds for alcohol consumption: combined analysis of individual-participant data for 599 912 current drinkers in 83 prospective studies](<https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(18)30134-X/fulltext>)

One of the reasons behind initializing this project by Angela M Wood et al. was the fact that there is no universal low-risk limit for alcohol consumption. In order to define this threshold - to see how much alcohol can be tolerated without causing significant harm - the researchers studied individual-participant data from 599912 current drinkers without previous cardiovascular disease.

The main finding was that the threshold for lowest risk for all-cause mortality was about 100 g per week. Long-term reduction of alcohol consumption (below the limit recommended in US guidelines) was associated with an increase of longer life expectancy. Exploratory analyses suggested that drinkers of beer or spirits, as well as binge drinkers, had the highest risk for all-cause mortality.

##### **Project 5:** [Consuming Alcohol in Moderation Can Lower Mortality Risks](https://www.acc.org/latest-in-cardiology/articles/2017/08/14/12/54/consuming-alcohol-in-moderation-can-lower-mortality-risks)

This is a study conducted by Bo Xi et al. and published in the Journal of the American College of Cardiology. As in the fourth project, heavy or binge drinking was associated with increased risk of all-cause mortality, while light and moderate alcohol intake might have, according to their study, a protective effect on all-cause and CVD-specific mortality in U.S. adults.

### Exploratory Data Analysis (**20 points**)

### Data Source: [National health and nutrition examination survey](https://www.kaggle.com/cdc/national-health-and-nutrition-examination-survey)

This data was obtained by the United States center of health statistics. It is an annual survey done for the civilian resident population of the United States. This is the public dataset corresponding to the period 2013-2014, and it was saved in the following folder: [/data/national-health-survey/](./data/national-health-survey/)

##### Data collection

The data collection of this survey was a complex task that involved several parts, starting with body measurements to questionnaires to evaluate the mental health. A further description of the methodology can be found [here](https://www.cdc.gov/nchs/data/series/sr_01/sr01_056.pdf).

#### Dataset description

The dataset is divided in several sections: demographic, medications or questionnaires. For our study, we are going to focus on the questionnaire section which is a dataset composed of several columns, where each one is a response to a certain question, such as: _In the past 12 months, how often did you drink any type of alcoholic beverage?_ The questionnaire dataset is composed of 10175 rows and 953 columns. 15 of those columns are the answers for questions related with alcohol consumption and 22 are related with mental health. Some of the features we considered during our analysis were the following:

- HighestEducationLevel
- MartialStatus
- Gender
- Age
- DifficultyConcentrating
- ThoughtSuicideLast2w
- MonthlyFamilyIncome
- FamilyPovertyIndex
- AlcoholAmountAvgPerMonth
- AlcoholAmountUnitPerMonth.

- Creates 5 well designed and formatted graphics (**15 points**, 3 each)

1. marital status average
2. gender average
3. a graph from mental health

  - The visual uses the appropriate visual encodings based on the data type (**1 point**)
  - Written interpretation of graphic is provided (**1 point**)
  - Clear axis labels, titles, and legends are included, where appropriate (**1 point**)

### Methods 

- **Strength of relationships**:

  - A formula describing how you believe your features (independent variables) are related to your outcome of interest (dependent variable) (**5 points**)
  I AM NOT SURE IF THIS IS WHAT KE ASKS FOR:
   We tried to asses the correlation between numerical features and average monthly alcohol consumption using the Pearson and Spearman coefficients.  
   We also made several graphs and ran some Chi Squared tests between selected categorical features and alcohol consumptions. 

  Based on our motivation for the project, we decided to include relevant features. It was rather challenging, since the dataset had a huge number of variables. During the analysis of the data, we looked at the distributions and planned to focus more on the variables that are normally distributed or uniformly distributed among all groups, however that was challenging since most of them were not. It was also important to have enough data, since in some cases, there were a lot of missing values.
 
  - Creating the appropriate model based on your dataset (**5 points**)
- **Prediction**: You must also make predictions for your outcome of interest. In doing so, you must demonstrate a clear use of:
  - Splitting your data into testing/training data (**2 points**)
  - Applying cross validation to your model (**3 points**)
  - Appropriately handling any missing values (**2 points**)
  - Appropriately using categorical variables (**3 points**)
  - Using a grid search to find the best parameters for you model of interest (**2 points**)
  - Employing the algorithm of interest (**3 points**)

### Results (**20 points**)
You must provide a clear interpretation of your statistical and machine learning results, including at least **one visual or table** for each.

- **Strengths of relationships**: For the features you included in your model, you must describe the strength (significance) and magnitude of the relationships. This can be presented in a table or chart, and pertinent observations should be described in the text.
(**10 points**)

Regarding the correlation between age and alcohol consumption, using the Pearson method gave us a 0.049 correlation, while the Spearman method returned a -0.145 coefficient. Although both coefficients are low, we notice a 10% difference between the two methods. This could possibly mean that there is a relation between these two variables that is not linear.


```{r, warning=FALSE}
cor(
  dataset$Age, 
  dataset$AlcoholAmountAvgPerMonth, 
  method = "pearson"
)
```
```{r, warning=FALSE}
cor(
  dataset$Age, 
  dataset$AlcoholAmountAvgPerMonth, 
  method = "spearman"
)
```

Trying to correlate monthly average alcohol consumption with the Poverty Index, using the Spearman method resulted in a 19% correlation. Although it is also quite low, we can assume a weak correlation, especially given the social context. 

```{r, warning=FALSE}
cor(
  income_dataset$FamilyPovertyIndex,
  income_dataset$AlcoholAmountAvgPerMonth,
  method = "spearman"
)
```


After computing the Chi squared tests, we concluded based on the low p-value that alcohol consumption depends on the gender and marital status. It was very clear from the graphs that the gender is related to alcohol consumption as well.

```{r, warning=FALSE}
chisq.test(dataset$Gender, dataset$AlcoholAmountAvgPerMonth, correct=FALSE)
```
```{r, warning=FALSE}
chisq.test(dataset$MaritalStatus, dataset$AlcoholAmountAvgPerMonth, correct=FALSE)
```

ADD VISUAL FOR GENDER


- **Predictions**: How well were you able to predict values in the dataset? You should both report appropriate metrics based on the type of outcome you're predicting (e.g., root mean squared error v.s. accuracy), as well as a high quality visual showing the strength of your model (**10 points**)

### Discussion and Future Work (**10 points**)

Based on _specific observations_ from the results section, the report clearly provides:

We were expecting to have more surprising and intriguing results, but unfortunately we were many times disappointed. Although we have chosen a huge dataset, it still felt like there was not enough data. Nevertheless, our methods had some clear results such as that alcohol consumption depends on the gender and marital status. Unfortuantely, we did not have the same sample size for all the marital status groups so this could be something to be improved for the future - a better dataset to check the dependence. 

The gender dependence was something we expected. In a new project this could futher be studied in relation to other variables, such as the amount of swear words used by each gender, crime activity and other factors that come in a full suite of cultural behaviour.

We also conclude that it is quite difficult to predict alcohol drinking habits as it is not strongly correlated to the factors about which we had information in the dataset. 

It is also important to note that the data was gathered from questionnaires, therefore its accuracy is not granted. For future, it would also be interesting to work with datasets that have a different source such as purchase statements.

Finally, as we improve our data processing skills and come across better data and in greater amounts, it would be ideal if we could use artificial neural networks to find patterns regarding alcohol consumption and mental health, as there migt be some tings that we are unable to detect using other methods. The results could potentially show the risks involved with alcohol consumption.
