---
title: "Alcohol consumption"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Include the main file in order to do preprocessing.
source("./main.R")
```

### Abstract

### Introduction and related work

CHANGE THAT USING AN SPECIFIC DATA DRIVEN QUESTION

We are interested to know which are the potential causes and consequences of alcohol consumption based on our data. Especially we looked for data related to young people. We planned on relating this topic with other domains such as education and health and find out if the causes of alcohol consumption is due to facts like the economical situation or Human Development Index of the countries.

We search for different related projects that could help us in how to drive our research:

##### **Project 1**: [How common is alcohol and drug dependency across the world?](https://ourworldindata.org/alcohol-and-drug-dependency)

This project from the website _Our World in Data_ presents the trends in alcohol consumption and substance use in countries all over the world, with data from 1990 to 2016. The aim is to give a better understanding of the global prevalence of substance use disorders.

Some of the conclusions in the research are:

- Across most countries, the number of alcohol use disorders is higher than other drug use disorders.
- Substance use disorders are more common among men than women.

##### **Project 2**: [SDR, selected alcohol-related causes, per 100 000](https://gateway.euro.who.int/en/indicators/hfa_293-1970-sdr-selected-alcohol-related-causes-per-100-000/)

This project carried out by the World Health Organization provide an interactive graph where we can see the age-standardized death rate per 100000 related to alcohol. The study includes causes such as cancer of esophagus and larynx, alcohol dependence syndrome, chronic liver disease, cirrhosis etc.

##### **Project 3:** [Alcohol as a risk factor for type 2 diabetes](https://care.diabetesjournals.org/content/32/11/2123.full-text.pdf)

This study by the University of Toronto tries to measure the relation between alcohol consumption and the presence of type 2 diabetes. For obtaining the data, the researchers have found several data sources both from public sources or from previous articles.

The main conclusion of this study is that a relationship exists between alcohol consumption and the presence or diabetes in an _u-shaped_ way, both for men and women.

##### **Project 4:** [Risk thresholds for alcohol consumption: combined analysis of individual-participant data for 599 912 current drinkers in 83 prospective studies](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(18)30134-X/fulltext)

One of the reasons behind initializing this project was the fact that there is no universal low-risk limit for alcohol consumption. In order to define this threshold - to see how much alcohol can be tolerated without causing significant harm -  the researchers studied individual-participant data from 599912 current drinkers without previous cardiovascular disease.

The main finding was that the threshold for lowest risk for all-cause mortality was about 100 g per week. Long-term reduction of alcohol consumption (below the limit recommended in US guidelines) was associated with an increase of longer life expectancy. Exploratory analyses suggested that drinkers of beer or spirits, as well as binge drinkers, had the highest risk for all-cause mortality.

##### **Project 5:** [Consuming Alcohol in Moderation Can Lower Mortality Risks](https://www.acc.org/latest-in-cardiology/articles/2017/08/14/12/54/consuming-alcohol-in-moderation-can-lower-mortality-risks)

As in the fourth project, heavy or binge drinking was associated with increased risk of all-cause mortality, while
light and moderate alcohol intake might have, according to their study, a protective effect on all-cause and CVD-specific mortality in U.S. adults. 

### Exploratory data analisys

```{r setup, include=FALSE}

ggplot(dataset, aes(x = HadLostConcientiousness)) + geom_histogram()

```




